---
title: "Explore association networks"
author: "Koen Van den Berge"
date: "9/9/2020"
output: 
  html_document:
    toc: true
    toc_float: true
---

Info:

  -  index from 1 to 6 stands for groups where: 1- HBC, 2-iOSN, 3-HBC*, 4-GBC, 5-mOSN, 6-all groups
  -  names with “dataTFtranscel” include only TF genes
  - all results received by performing zinb1 algorithm with alpha(significance levels tests)  $\alpha=2\times\text{pnorm}(n^{0.15},\text{lower.tail=F})$

```{r}
library(igraph)
library(rgl)
library(rafalib)
library(slingshot)
library(msigdbr)
library(fgsea)
library(knitr)
library(leiden)
sds <- readRDS("sling.rds")
clDatta <- readRDS("dattaCl_noResp_noMV.rds")
cols <- c("#E41A1C", "#377EB8", "#4DAF4A", "#984EA3", "#FF7F00", "#FFFF33", "#A65628", "#F781BF", "#999999")
view3d( theta = 10, phi = 10)
# Datta labels
plot3d(reducedDim(sds), aspect = 'iso', col=cols[clDatta], alpha=.6, xlab="UMAP1", ylab="UMAP2", zlab="UMAP3", box=FALSE, top=TRUE)
load("list_TF.RData")
```

# Transcription factor networks

## Neuronal lineage

HBC* - GBC - iOSN - mOSN

```{r}
## activated HBC
load("3-dataTFtranscelsalpha059515result.RData")
adjHBCAct <- adj.zinb1
dimnames(adjHBCAct) <- list(colnames(datatest), colnames(datatest))
adjHBCAct <- adjHBCAct[!rowSums(adjHBCAct) == 0,]
adjHBCAct <- adjHBCAct[,!colSums(adjHBCAct) == 0]
hub.HBCact <- which(apply(adjHBCAct,2,sum)>=9)
rm(adj.zinb1, datatest)


## GBC
load("4-dataTFtranscelsalpha059515result.RData")
adjGBC <- adj.zinb1
dimnames(adjGBC) <- list(colnames(datatest), colnames(datatest))
adjGBC <- adjGBC[!rowSums(adjGBC) == 0,]
adjGBC <- adjGBC[,!colSums(adjGBC) == 0]
hub.GBC <- which(apply(adjGBC,2,sum)>=9)
rm(adj.zinb1, datatest)


## iOSN
load("2-dataTFtranscelsalpha059515result.RData")
adjiOSN <- adj.zinb1
dimnames(adjiOSN) <- list(colnames(datatest), colnames(datatest))
adjiOSN <- adjiOSN[!rowSums(adjiOSN) == 0,]
adjiOSN <- adjiOSN[,!colSums(adjiOSN) == 0]
hub.iOSN <- which(apply(adjiOSN,2,sum)>=9)
rm(adj.zinb1, datatest)


## mOSN
load("5-dataTFtranscelsalpha059515result.RData")
adjmOSN <- adj.zinb1
dimnames(adjmOSN) <- list(colnames(datatest), colnames(datatest))
adjmOSN <- adjmOSN[!rowSums(adjmOSN) == 0,]
adjmOSN <- adjmOSN[,!colSums(adjmOSN) == 0]
hub.mOSN <- which(apply(adjmOSN,2,sum)>=9)
rm(adj.zinb1, datatest)

saveRDS(list("hub.hbcAct" = hub.HBCAct, 
             "hub.GBC" = hub.GBC, 
             "hub.iOSN" = hub.iOSN, 
             "hub.mOSN" = hub.mOSN), file="hub_nodes.rds")

```



```{r}
degreeDf <- data.frame(degree=1:52,
                       hbcAct=NA,
                       gbc=NA,
                       iOSN=NA,
                       mOSN=NA)
degreeHBCAct <- table(rowSums(adjHBCAct))
degreeDf$hbcAct[match(names(degreeHBCAct), degreeDf$degree)] <- degreeHBCAct
degreeGBC <- table(rowSums(adjGBC))
degreeDf$gbc[match(names(degreeGBC), degreeDf$degree)] <- degreeGBC
degreeiOSN <- table(rowSums(adjiOSN))
degreeDf$iOSN[match(names(degreeiOSN), degreeDf$degree)] <- degreeiOSN
degreemOSN <- table(rowSums(adjmOSN))
degreeDf$mOSN[match(names(degreemOSN), degreeDf$degree)] <- degreemOSN


mypar(mfrow=c(2,2))
barplot(degreeDf$hbcAct, names=degreeDf$degree, main="HBC*: Node-level degree", ylim=c(0,400))
barplot(degreeDf$gbc, names=degreeDf$degree, main="GBC: Node-level degree", ylim=c(0,400))
barplot(degreeDf$iOSN, names=degreeDf$degree, main="iOSN: Node-level degree", ylim=c(0,400))
barplot(degreeDf$mOSN, names=degreeDf$degree, main="mOSN: Node-level degree", ylim=c(0,400))
```

## Extract 2-core of the network

```{r}
extractCore <- function(A, degree = 3){
  converg <- FALSE
  old.nrow <- nrow(A)
  while(!converg){
    d <- colSums(A)
    to.keep <- which(d>=degree)
    if(old.nrow==length(to.keep)){
      converg <- TRUE
    }
    old.nrow <- length(to.keep)
    A <- A[to.keep,to.keep]
  }
  return(A)
}

adjHBCAct2 <- extractCore(adjHBCAct, degree = 2)
adjGBC2 <- extractCore(adjGBC, degree = 2)
adjiOSN2 <- extractCore(adjiOSN, degree = 2)
adjmOSN2 <- extractCore(adjmOSN, degree = 2)

mypar(mfrow=c(2,2))
gHBCAct2 <- graph_from_adjacency_matrix(adjHBCAct2,mode="undirected")
plot(gHBCAct2,vertex.size=3,vertex.label=NA,edge.width=0.6, main="HBC*")
gGBC2 <- graph_from_adjacency_matrix(adjGBC2,mode="undirected")
plot(gGBC2,vertex.size=3,vertex.label=NA,edge.width=0.6, main="GBC")
giOSN2 <- graph_from_adjacency_matrix(adjiOSN2,mode="undirected")
plot(gGBC2,vertex.size=3,vertex.label=NA,edge.width=0.6, main="iOSN")
gmOSN2 <- graph_from_adjacency_matrix(adjmOSN2,mode="undirected")
plot(gGBC2,vertex.size=3,vertex.label=NA,edge.width=0.6, main="mOSN")

saveRDS(list("hbcAct" = adjHBCAct2, 
             "GBC" = adjGBC2, 
             "iOSN" = adjiOSN2, 
             "mOSN" = adjmOSN2), file="adjMatrices_2core.rds")
```

## Community detection

```{r}
library(RColorBrewer)
library(HCD)

gradualPlot <- function(cl, A){
  ncl <- length(unique(cl))
  for(cc in 1:ncl){
    id <- which(cl %in% (1:cc))
    curG <- graph_from_adjacency_matrix(A[id,id],mode="undirected")
      V(curG)$color <- cols[cl[names(V(curG))]]
      lo <- layout_with_fr(curG)
     plot(curG,
         vertex.size=3,
       vertex.label=NA,
       edge.width=0.6,
       vertex.frame.color=cols[cl[id]],
       color=V(curG)$color,
       layout = lo
       )
  }
}

## C5 category is according to gene ontology grouping: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4707969/pdf/nihms-743907.pdf
geneSets <- msigdbr(species = "Mus musculus", category = "C5", subcategory = "BP")

gsea <- function(genes, background, geneSets, n=10, minSize=5, name=NULL){
  ### filter background to only include genes that we assessed.
  geneSets <- geneSets[geneSets$gene_symbol %in% background,]
  m_list <- geneSets %>% split(x = .$gene_symbol, f = .$gs_name)
  # gene set must have at least minSize genes in background.
  m_list <- m_list[unlist(lapply(m_list, length)) >= minSize]
  
  overlapPval <- unlist(lapply(m_list, function(gs){
    # genes in community and gene set
    inBoth <- sum(genes %in% gs)
    # genes in community and not in gene set
    inComOnly <- length(genes) - inBoth
    # genes in background and gene set
    inGsBack <- sum(background %in% gs)
    # genes in background and not in gene set
    outGsBack <- length(background) - inGsBack
    m <- matrix(c(inBoth, inComOnly,
           inGsBack, outGsBack),
           nrow =2, ncol=2, byrow=TRUE,
           dimnames = list(c("in community", "out community"),
                           c("in gene set", "out gene set")))
    fis <- fisher.test(m, alternative = "greater")
    pval <- fis$p.value
    return(pval)
  }))
  padj <- p.adjust(overlapPval, "fdr")
  oo <- order(overlapPval, decreasing=FALSE)
  res <- data.frame(geneSet = names(m_list)[oo[1:n]],
                    pval = overlapPval[oo[1:n]],
                    padj = padj[oo[1:n]],
                    row.names = NULL)
  kable(res, caption=name, label=name)
}


```

### HBC*

```{r}
#set.seed(2)
clHBCAct <- leiden(adjHBCAct2, resolution_parameter=.4, seed = 7)
table(clHBCAct)

gHBCAct <- graph_from_adjacency_matrix(adjHBCAct2,mode="undirected")
cols <- brewer.pal(8, "Dark2")
V(gHBCAct)$color <- cols[clHBCAct[names(V(gHBCAct))]]
plot(gHBCAct,
     vertex.size=3,
     vertex.label=NA,
     #vertex.frame.color=V(gHBC)$color,
     color=V(gHBCAct)$color,
     edge.width=0.6)
comHBCAct <- make_clusters(gHBCAct, membership = clHBCAct)
plot(comHBCAct, gHBCAct,vertex.size=3, vertex.label=NA, edge.width=0.6)

## gradual plot
gradualPlot(clHBCAct, adjHBCAct2)


## gene set enrichment on HCD results
ncl <- length(unique(comHBCAct$membership))
kab <- list()
# kabNames <- c("Krt / cell cycle / stimulus",
#                 "Chromatin",
#                 "Epigenetics, cell cycle",
#                 "Cell cycle")
kabNames <- c("Cell cycle, DNA damage",
                "Epigenetics",
                "(Epithelial) cell differentiation",
                "Cell cycle, DNA replication",
              "DNA replication")
for(kk in 1:ncl){
  genes <- rownames(adjHBCAct2)[which(comHBCAct$membership == kk)]
  kab[[kk]] <- gsea(genes = genes,
       background = TF.list,
       geneSets = geneSets,
       name = kabNames[kk])
}
kab
```


### GBC

```{r}
set.seed(3)
clGBC <- leiden(adjGBC2, resolution_parameter=.4, seed=7)
table(clGBC)

gGBC <- graph_from_adjacency_matrix(adjGBC2,mode="undirected")
cols <- brewer.pal(8, "Dark2")
V(gGBC)$color <- cols[clGBC[names(V(gGBC))]]
plot(gGBC,
     vertex.size=3,
     vertex.label=NA,
     #vertex.frame.color=V(gHBC)$color,
     color=V(gGBC)$color,
     edge.width=0.6)
comGBC <- make_clusters(gGBC, membership = clGBC)
plot(comGBC, gGBC,vertex.size=3, vertex.label=NA, edge.width=0.6)

## gradual plot
gradualPlot(clGBC, adjGBC2)

## gene set enrichment 
ncl <- length(unique(comGBC$membership))
kab <- list()
# kabNames <- c("Chromatin, neuron/cell development",
#                 "P53, DNA replication, cell cycle",
#                 "Cell differentiation / signaling",
#                 "Notch, signaling",
#               "DNA repair")
kabNames <- c("P53, DNA replication",
                "Notch signaling",
                "Cell differentiation / signaling",
                "RNA splicing, DNA repair, epigenetics",
                "NIK/NF Kappa-B signaling")
for(kk in 1:ncl){
  genes <- rownames(adjGBC2)[which(comGBC$membership == kk)]
  print(gsea(genes = genes,
       background = TF.list,
       geneSets = geneSets,
       name = kabNames[kk]))
}
kab
```



### iOSN

```{r}
set.seed(4)
cliOSN <- leiden(adjiOSN2, resolution_parameter=.3, seed=7)
table(cliOSN)


giOSN <- graph_from_adjacency_matrix(adjiOSN2,mode="undirected")
cols <- brewer.pal(8, "Dark2")
V(giOSN)$color <- cols[cliOSN[names(V(giOSN))]]
plot(giOSN,
     vertex.size=3,
     vertex.label=NA,
     color=V(giOSN)$color,
     edge.width=0.6)
# igraph plot
comiOSN <- make_clusters(giOSN, membership = cliOSN)
plot(comiOSN, giOSN,vertex.size=3, vertex.label=NA, edge.width=0.6)

## gradual plot
gradualPlot(cliOSN, adjiOSN2)

## gene set enrichment on HCD results
ncl <- length(unique(comiOSN$membership))
kab <- list()
kabNames <- c("Neurogenesis",
              "Defense (virus) response",
              "P53, membrane",
              "Axon extension, cytokinesis, epigenetic gene expression regulation",
              "Cytokinesis, DNA replication",
              "Interleukins, cell cycle, protein localization",
              "Stem cell population maintenance, transcription")
for(kk in 1:ncl){
  genes <- rownames(adjiOSN2)[which(comiOSN$membership == kk)]
  print(gsea(genes = genes,
       background = TF.list,
       geneSets = geneSets,
       name = kabNames[kk]))
}
kab
```



### mOSN

```{r}
set.seed(2)
clmOSN <- leiden(adjmOSN2, resolution_parameter=.4, seed=7)
table(clmOSN)

gmOSN <- graph_from_adjacency_matrix(adjmOSN2,mode="undirected")
cols <- brewer.pal(8, "Dark2")
V(gmOSN)$color <- cols[clmOSN[names(V(gmOSN))]]
plot(gmOSN,
     vertex.size=3,
     vertex.label=NA,
     color=V(gmOSN)$color,
     edge.width=0.6)
# igraph plot
commOSN <- make_clusters(gmOSN, membership = clmOSN)
plot(commOSN, gmOSN,vertex.size=3, vertex.label=NA, edge.width=0.6)

## gradual plot
gradualPlot(clmOSN, adjmOSN2)

## gene set enrichment on HCD results
ncl <- length(unique(commOSN$membership))
kab <- list()
kabNames <- c("Chromatin",
              "Chromatin, transcription",
              "TGBF-Beta pathway, neuron death",
              "Viral processes",
              "Cell projection, others")
for(kk in 1:ncl){
  genes <- rownames(adjmOSN2)[which(commOSN$membership == kk)]
  print(gsea(genes = genes,
       background = TF.list,
       geneSets = geneSets,
       name = kabNames[kk]))
}
kab
```
